---
title: "Sensitivity analysis"
output: html_document
date: "2025-04-04"
---

```{r, echo = FALSE}
require(mmil)
require(glmnet)
require(ggplot2)
require(rsample)
require(data.table)
require(pROC)

set.seed(2345)
```

# Background

# How to perform a sensitivity analysis

We start by simulating data, which we can do with the `mmil_simulate_data` function in the **mmil** package.

Here, we simulate 1000 cells with 100 features each, and rho = 0.9. The argument `shift` = 0.4 allows us to control the strength of the signal. In this case, we add 0.4 to the first 50 covariates in disease-associated cells.

```{r}
p = 100
n = 1000
zeta = 0.5 # (default in mmil_simulate_data)
shift = 1
data = mmil_simulate_data(num_cells = n, 
                          rho = 0.9,
                          num_features = p,
                          shift = shift)

data.test = mmil_simulate_data(num_cells = n, 
                          rho = 0.9,
                          num_features = p,
                          shift = shift)
```

Here, we'll try a wide range of values: $\rho = 0.10, 0.25, 0.50, 0.75, 0.90$. In real settings, we should consider values in a neighborhood that we think is reasonable, given what we know about our problem.

```{r}
rhos = c(.1, .25, .5, .75, .9)
```

We also need to choose a set of $\lambda$ values to consider.
```{r}
lambda_path = c(1e-3, 5e-4)
```

We set our case control adjustment to 0, because we know that our train and test distributions are the same.

```{r}
case.control.adj <- 0
```

Now, we have two parameters to choose: (1) the $\lambda$ parameter (for lasso) and (2) the $\rho$ parameter (for MMIL). We will choose these parameters using cross validation: for each ($\lambda$, $\rho$), we will compute and store the CV log likelihood. Then, for each $\rho$, choose lambda that optimizes the CV log likelihood. We will fit a model for each of these ($\rho$, $\lambda$) pairs and compare their CV log likelihoods.

First, we make our CV folds:

```{r}
# CV to choose lambda (glmnet parameter)
n.folds = 3
cv_folds <- vfold_cv(
    data.frame(X = data$X,
               z = data$z),
    v = n.folds)
```
**Note**: We simulated our data so that all cells are independent, but in real-world settings, remember to ensure that cross-validation folds respect person-level grouping—that is, all cells from a given individual should be assigned to the same fold. This can be done with the function `group_vfold_cv`.


And now we model:
```{r}
results = rep(NULL, 3)
for(lambda in lambda_path){
    for(rho_guess in rhos){
      print(c(rho_guess, lambda))
    
      mean.ll = 0
      for (i in 1:n.folds) {
        train_data <- analysis(cv_folds$splits[[i]])  
        test_data  <- assessment(cv_folds$splits[[i]])

        mixture.model <- mmil_fit_glmnet(as.matrix(train_data[, 1:p]),
                                         train_data[, "z"],
                                         rho = .9,
                                         zeta = zeta,
                                         lambda = lambda,
                                         case_control_adjustment = case.control.adj,
                                         num_iterations = 50,
                                         early_stopping_patience = 3,
                                         fit_entire_glmnet_path = FALSE)

        
        preds <- predict(mixture.model, as.matrix(test_data[, 1:p]), s=lambda)
        ll <- mmil_loglik(test_data[, "z"], preds,
                          rho = rho_guess,
                          num_cells_disease = sum(train_data[, "z"]),
                          num_cells = nrow(train_data)
                          )
        mean.ll = mean.ll + ll/n.folds
      }
      results = rbind(results, c(rho_guess, lambda, mean.ll))
    }
}
colnames(results) = c("rho", "lambda", "ll")
results <- data.table(results[2:nrow(results), ])
```

Here is our performance:
```{r}
best_lambdas <- results[, .(ll = max(ll), lambda), by = rho]

ggplot(best_lambdas) + 
  geom_point(aes(x=rho, y=ll)) + 
  geom_line(aes(x=rho, y=ll)) +
  labs(x = expression(rho), y = "Observed log likelihood",
       title = "Observed cross-validation log likelihood",
       subtitle = expression("for different values of "*rho))
```

```{r}
cat("Best rho:", results[which.max(ll), rho])
```

This analysis tells us that we should select $\rho = 0.75$ -- the value that gives us the largest log likelihood. This is pretty close to the *true* $\rho$ that we used to simulate data. We now fit our final model using the full dataset:
```{r}
# Fit the final model:
rho = results[which.max(ll), rho]
lambda = results[which.max(ll), lambda]
mixture.model    <- mmil_fit_glmnet(data$X, 
                                        data$z,
                                        rho = rho,
                                        zeta = zeta, 
                                        lambda = lambda,
                                        case_control_adjustment = case.control.adj,
                                        num_iterations = 100,
                                        early_stopping_patience = 3)
```
 

We can predict for our test data:
```{r}
preds <- as.vector(predict(mixture.model, newx = data.test$X, s=lambda, type = "response"))
head(preds)
```

And compute the AUROC -- though typically, the AUROC is unknowable (because the $y$ labels are unknown): 
```{r}
roc_obj <- roc(data.test$true_y, preds, 
               direction = "<",
               levels = c(0, 1))  
auc(roc_obj)
```


